Scheduling and Vectorization for MPC

### 2022-03-01 paper_SIMD.tex
wip on problem statement and costs

\usepackage[utf8]{inputenc}
\usepackage[american]{babel}
\usepackage[backend=biber,style=ACM-Reference-Format]{biblatex}
\addbibresource{../library.bib}
\addbibresource{../cryptobib/abbrev2.bib}
\addbibresource{../cryptobib/crypto_crossref.bib}

@@

@@

We have def-use dependency graph $G(V, E)$ for the input program. Each vertex/node in this graph represents an operation/instruction in the program and an edge represents dependence i.e. an edge from $V_1$ to $V_2$ means the result of operation in $V_1$ is used in the operation in $V_2$. To build a parallel schedule, we translate this graph $G(V, E)$ into another dependency graph $G'(V', E')$ such that:

\begin{enumerate}
    \item We collapse multiple vertices of the same type (i.e. same operation) in $V$ into a single vertex in $V'$.
    \item dependencies are not broken i.e. If there is an edge from $V_1$ to $V_2$ ($V_1, V_2 \in V$), and we put them in vertices $V'_1, V'_2 \in G'$ respectively, then an edge must exist from $V'_1$ to $V'_2$.
\end{enumerate}

We want to build a graph $G'$ with fewest possible vertices since fewer vertices means more parallelization. \ishaq{TODO: make it more concrete, pick some variables e.g. $\alpha, \beta$ like above and show 1) parallel schedule will have lower cost (this should be easy). and 2) some sort of optimality? this would be hard}

@@

\ana{Ishaq?}\ishaq{TODO: As a first step, pull the super-sequence proof from Lindsay's thesis? }

@@

\COMMENT { Phase 3: Remove unnecessary dimensionality. }

@@

\subsection{Extension of Basic Vectorization}

@@

{\small
\begin{verbatim}
for i in range(N):
  A[i] = B[i] + 10;
  B[i] = A[i] * D[i-1];
  C[i] = A[i] * D[i-1];
  D[i] = B[i] * C[i];
\end{verbatim}
}

@@

{\small
\begin{verbatim}
for i in range(N):
1. A_1 = PHI(A_0,A_2)
2. B_1 = PHI(B_0,B_2)
3. C_1 = PHI(C_0,C_2)
4. D_1 = PHI(D_0,D_2)
5. A_2 = update(A_1, i, B_1[i] + 10); 
6. B_2 = update(B_1, i, A_2[i] * D_1[i-1]);
7. C_2 = update(C_1, i, A_2[i] * D_1[i-1]);
8. D_2 = update(D_1, i, B_2[i] * C_2[i]);
\end{verbatim}
}

\begin{comment}
\begin{algorithmic}
\STATE for i in range(N):
\STATE \hspace{0.25cm} A[i] = B[i] + 10;
\STATE \hspace{0.25cm} B[i] = A[i] * D[i-1];
\STATE \hspace{0.25cm} C[i] = A[i] * D[i-1];
\STATE \hspace{0.25cm} D[i] = B[i] * C[i];
\end{algorithmic}

\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{A}_0$ = $\phi(\texttt{A},\texttt{A}_1)$
\STATE 2. $\texttt{B}_0$ = $\phi(\texttt{B},\texttt{B}_1)$
\STATE 3. $\texttt{C}_0$ = $\phi(\texttt{C},\texttt{C}_1)$
\STATE 4. $\texttt{D}_0$ = $\phi(\texttt{D},\texttt{D}_1)$
\STATE 5. $\texttt{A}_1$ = $\texttt{A}_0$; $\texttt{A}_1$[i] = $\texttt{B}_0$[i] + 10; \COMMENT{equiv. to Update}
\STATE 6. $\texttt{B}_1$ = $\texttt{B}_0$; $\texttt{B}_1$[i] = $\texttt{A}_1$[i] * $\texttt{D}_0$[i-1];
\STATE 7. $\texttt{C}_1$ = $\texttt{C}_0$; $\texttt{C}_1$[i] = $\texttt{A}_1$[i] * $\texttt{D}_0$[i-1];
\STATE 8. $\texttt{D}_1$ = $\texttt{D}_0$; $\texttt{D}_1$[i] = $\texttt{B}_1$[i] * $\texttt{C}_1$[i];
\end{algorithmic}
\end{comment}

There is a cycle around \texttt{B\_1 = PHI(B\_0,B\_2)} that includes statement 
\texttt{A\_1 = update(A\_0, i, B\_1[i] + 10)} and that statement won't be vectorized even 
though in fact there is no loop-carried dependency from the write of \texttt{B\_1[i]} at 6 to the
read of \texttt{ ... = B\_1[i]} at 8.

@@

Consider the earlier example. There is a single loop, $i$. Clearly, there is no pair $\underline{i}$ and $\underline{i}'$, where $\underline{i} < \underline{i}'$ that make $\underline{i} = \underline{i}'$ 
due to the def-use pairs of \texttt{A} 5-6 and 5-7.
Therefore, we remove the back edge from 5 to the phi-node 1. Analogously, we remove the back edges from 6 to 2 and from 7 to 3. However, there are many values $\underline{i} < \underline{i'}$ that make $\underline{i} = \underline{i'}-1$ and the back edge from 8 to 4 remains (def-use pairs for \texttt{D}). As a result of removing these spurious edges, Basic Vectorization will find that statement 5 is vectorizable. Statements 6, 7 and 8 will correctly appear in the FOR loop.

Note however, that this step renders some array phi-nodes target-less. We handle target-less phi-nodes with a minor extension of Basic Vectorization. 
First, we merge closures that update the same array. This simplifies handling of array $\phi$-nodes: if each closure is turned into a separate loop
each loop will need to have its own array phi-node to account for the update and this would complicate the analysis.
Second, we add the target-less node of array \texttt{A} back to the closure that updates \texttt{A} ---
the intuition is, even if there is no loop-carried dependence from writes to reads on \texttt{A}, \texttt{A} is written and the write (i.e., update) cannot be vectorized; 
therefore, the updated array has to carry to the next iteration of the loop. Third, in cases when the phi-node remains target-less, i.e., cases when the
array write can be vectorized, we have to properly remove the phi-node replacing uses of the left-hand side of the phi-node with its arguments.

@@

\ana{TODO: WIP, get to BV without MUX refinement for now? For now, skip MUX refinement.}

@@

@@

For now, we restrict array updates to \emph{canonical updates}. Assume (for simplicity) a two-dimensional array \texttt{A[I,J]}. A canonical update is the following:
\begin{verbatim}
for i in range(I):
  for j in range(J):
    ...
    A[i,j] = ...
    ...
\end{verbatim}

The update \texttt{A[i,j]} can be nested into an inner loop and there may be multiple updates, i.e., writes to \texttt{A[i,j]}. However, update such as \texttt{A[i-1,j] = ...} or \texttt{A[i-1,j-1] = ...}, etc., is not allowed. Additionally, while there could be several different loops that perform canonical updates, they must be of the same dimensionality, i.e., an update of higher or lower dimension, e.g., \texttt{A[i,j,k] = ...} is not allowed. We compute the \emph{canonical dimensionality} of each write array by examining the array writes in the original program and rejecting programs that violate the canonical write restriction. This restriction simplifies reasoning in this early stage of the compiler; we will look to relax the restriction in future work.

@@

Reads through an arbitrary formula, such as \texttt{A[i-1]} for example, are allowed; currently, the projection function returns dummy values if the read formula is out of bounds; we assume the programmer ensures that the program still computes correct output in this case.

@@

\begin{comment}
There are two changes to Basic Vectorization to account for write arrays. We can account for them during processing of raise dimension/drop dimension in Phase 1 of Basic Vectorization.

One change to Basic vectorization is the expansion of dimension if the array write or read occurs in a nested loop. That is, if there is an update $\texttt{A}[i,j] = ...$ that occurs in loop dimensionality $i,j,k$, $\texttt{A}[i,j]$ will be rewritten into $\texttt{A}[i,j,k]$. Similarly, a read $\texttt{A}[f(i,j)]$ will be rewritten into $\texttt{A}[f(i,j),k]$. ($\texttt{A}[f(i,j),k]$ will be reshaped as $f(i,j)*I*J+k$ during Phase 1 of Basic Vectorization. \ana{TODO: Double check.}) 
\end{comment}

In addition to the changes for the handling of target-less phi-nodes, Basic Vectorization has to handle differently def-use edges $X \rightarrow Y$ where $X$ defines and $Y$ uses an array variable. The definition can be an update \texttt{A\_2 = update(A\_1,i,...)} or a pseudo $\phi$-node \texttt{A\_2 = PHI(A\_0,A\_1)}, etc.. Note that $\phi$-nodes for arrays have no subscript operations the way there are subscript operations in analysis-introduced arrays representing scalars. In addition, array variables are different from scalar variables. Raising and dropping dimension still apply, however, the dimension cannot be raised or dropped to a dimension lower than the canonical one:

@@

First, the canonical dimensionality of all \texttt{A,B,C} and \texttt{D} is 1. 
After Phase 1 of Basic Vectorization the Aiken's array write example will be (roughly) as follows:

{\small
\begin{verbatim}
for i in range(N):
1. A_1 = PHI(A_0,A_2)
2. B_1 = PHI(B_0,B_2)
3. C_1 = PHI(C_0,C_2)
4. D_1 = PHI(D_0,D_2)
5. A_2 = update(A_1, I, B_1[I] + 10); 
6. B_2 = update(B_1, I, A_2[I] * D_1[I-1]);
7. C_2 = update(C_1, I, A_2[I] * D_1[I-1]);
8. D_2 = update(D_1, I, B_2[I] * C_2[I]);
\end{verbatim}
}

Note that since all def-uses are same-level (i.e., reads and writes of the array elements) 
no raise dimension or drop dimension happens.

\begin{comment}
\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{A}_1$ = $\phi(\texttt{A}_0,\texttt{A}_2)$
\STATE 2. $\texttt{B}_1$ = $\phi(\texttt{B}_0,\texttt{B}_2)$
\STATE 3. $\texttt{C}_1$ = $\phi(\texttt{C}_0,\texttt{C}_2)$
\STATE 4. $\texttt{D}_1$ = $\phi(\texttt{D}_0,\texttt{D}_2)$
\STATE 5. $\texttt{A}_2 = update(\texttt{A}_1, \stackrel{\rightarrow}{i}, \texttt{B}_1[\stackrel{\rightarrow}{i}] + 10)$;
\STATE 6. $\texttt{B}_2 = update(\texttt{B}_1, \stackrel{\rightarrow}{i}, \texttt{A}_2[\stackrel{\rightarrow}{i}] * \texttt{D}_1[\stackrel{\rightarrow}{i-1}])$;
\STATE 7. $\texttt{C}_2 = update(\texttt{C}_1, \stackrel{\rightarrow}{i}, \texttt{A}_2[\stackrel{\rightarrow}{i}] * \texttt{D}_1[\stackrel{\rightarrow}{i-1}])$;
\STATE 8. $\texttt{D}_2 = update(\texttt{D}_1; \stackrel{\rightarrow}{i}, \texttt{B}_2[\stackrel{\rightarrow}{i}] * \texttt{C}_2[\stackrel{\rightarrow}{i}])$;
\end{algorithmic}
\end{comment}

Phase 2 computes the closure of 4; $cl = \{4,6,7, 8\}$ 
while 5 is vectorizable. Recall that 1,2, and 3 are target-less phi-nodes. Since the closure $cl$ includes updates to \texttt{B} and \texttt{C},
the corresponding phi-nodes are added back to the closure and the def-use edges are added back to the target-less nodes. The uses of \texttt{A\_1} and \texttt{B\_1} in the vectorized statement
turn into uses of \texttt{A\_0} and \texttt{B\_0} respectively; this is done for all original target-less phi-node. (But note that \texttt{A\_0} is irrelevant; the update writes into array \texttt{A\_2} in parallel.)
Finally, the target-less phi-node for \texttt{A} is discarded.

{\small
\begin{verbatim}
1. A_2 = update(A_0, I, ADD_SIMD(B_0[I],10));
   equiv. to A_2[I] = ADD_SIMD(B_0[I],10)
FOR i=0; i<N; i++; // MOTION loop
2. B_1 = PHI(B_0,B_2)
3. C_1 = PHI(C_0,C_2)
4. D_1 = PHI(D_0,D_2)
5. B_2 = update(B_1, i, A_2[i] * D_1[i-1]);
   equiv. to B_2 = B_1; B_2[i] = A_2[i] * D_1[i-1];
6. C_2 = update(C_1, i, A_2[i] * D_1[i-1]);
7. D_2 = update(D_1, i, B_2[i] * C_2[i]);
\end{verbatim}
}

\begin{comment}
\begin{algorithmic}
\STATE 1. $\texttt{A}_2 = update(\texttt{A}_0; \stackrel{\rightarrow}{i}, \mathit{ADD\_SIMD}(\texttt{B}_0[\stackrel{\rightarrow}i],[10,...])$ \COMMENT{Fully vectorized, size N.}
\STATE FOR i=0; i<N; i++; \COMMENT{ MOTION loop }
\STATE 2. $\texttt{B}_1$ = $\phi(\texttt{B}_0,\texttt{B}_2)$
\STATE 3. $\texttt{C}_1$ = $\phi(\texttt{C}_0,\texttt{C}_2)$
\STATE 4. $\texttt{D}_1$ = $\phi(\texttt{D}_0,\texttt{D}_2)$
\STATE 5. $\texttt{B}_2 = update(\texttt{B}_1, i, \mathit{MUL}(\texttt{A}_2[i], \texttt{D}_1[{i-1}])$
\STATE 6. $\texttt{C}_2 = update(\texttt{C}_1, i, \mathit{MUL}(\texttt{A}_2[i], \texttt{D}_1[{i-1}])$
\STATE 7. $\texttt{D}_2 = update(\texttt{D}_1, i, \mathit{MUL}(\texttt{B}_2[i], \texttt{C}_1[i])$
\end{algorithmic}
\end{comment}

\paragraph{Example 2} Now consider the MPC Source of Histogram:

{\small
\begin{verbatim}
for i in range(0, num_bins):
   res1 = PHI(res, res2)
   for j in range(0, N):
      res2 = PHI(res1, res3)
      tmp1 = (A[j] == i)
      tmp2 = (res2[i] + B[j])
      tmp3 = MUX(tmp1, res2[i], tmp2)
      res3 = Update(res2, i, tmp3)     
return res1
\end{verbatim}
}

The canonical dimensionality of \texttt{res} is 1. Also, the phi-node \texttt{res2 = PHI(res1, res3)} is a target-less phi-node (the implication being that the inner for loop can be vectorized across $i$). After Phase 1, Basic vectorization produces the following (everything is implicitly vectorized along $i$ and $j$ and the update is turned into a regular write):

{\small
\begin{verbatim}
A1 = raise_dim(A, j, ((i:num_bins),(j:N)))
B1 = raise_dim(B, j, ((i:num_bins),(j:N)))
I = raise_dim(i, ((i:num_bins),(j:N)))
for i in range(0, num_bins):
   res1 = PHI(res, res2) // target-less, no drop_dim
   res1' = raise_dim(res1, (j:N))
   for j in range(0, N):
      res2 = PHI(res1', res3)
      tmp1 = (A1 == I)
      tmp2 = (res2 + B1)
      tmp3 = MUX(tmp1, res2, tmp2)
      res3 = res2 // copy propagation: res2 into res3
      res3 = tmp3 // "normal": res3[I,J] = tmp3[I,J]    
return res1
\end{verbatim}
}

Processing the inner loop in Phase 2 vectorizes \texttt{tmp1 = (A1 == I)} along the $j$ dimension but leaves the rest of the statements in a MOTION loop. Processing the outer loop is interesting. This is because the PHI node is a target-less node, and therefore, there are no closures! Thus everything can be vectorized along the $i$ dimension:

{\small
\begin{verbatim}
A1 = raise_dim(A, j, ((i:num_bins),(j:N)))
B1 = raise_dim(B, j, ((i:num_bins),(j:N)))
I1 = raise_dim(i, ((i:num_bins),(j:N)))

tmp1[I,J] = (A1[I,J] == I1[I,J])

res1' = raise_dim(res, (j:N)) // raising res to a 2-dim. array
for j in range(0, N):
   res2 = PHI(res1', res3)
   tmp2[I,j] = (res2[I,j] + B1[I,j])
   tmp3[I,j] = MUX(tmp1[I,j], res2[I,j], tmp2[I,j])
   res3 = res2
   res3[I,j] = tmp3[I,j]
res1 = drop_dim(res2) // Have to add drop_dim before removing redundant phi-node      
return res1
\end{verbatim}
}

The cleanup phase can then get rid of $j$ in the res variables and tmp variables, so the loop operates over the num\_bins-size vectors. 
This will lead to dropping raise\_dim and drop\_dim as well:

{\small
\begin{verbatim}
A1 = raise_dim(A, j, ((i:num_bins),(j:N)))
B1 = raise_dim(B, j, ((i:num_bins),(j:N)))
I1 = raise_dim(i, ((i:num_bins),(j:N)))

tmp1[I,J] = (A1[I,J] == I1[I,J])

for j in range(0, N):
   res2 = PHI(res1', res3) // now 1-dim array
   tmp2[I] = (res2[I] + B1[I,j])
   tmp3[I] = MUX(tmp1[I,j], res2[I], tmp2[I])
   res3 = res2
   res3[I] = tmp3[I]
return res2
\end{verbatim}
}

>>>>

\usepackage[utf8]{inputenc}
\usepackage[american]{babel}
\usepackage[backend=biber,style=ACM-Reference-Format]{biblatex}
\addbibresource{library.bib}
\addbibresource{../cryptobib/abbrev2.bib}
\addbibresource{../cryptobib/crypto_crossref.bib}

@@

% for references, use it like this: \cref{some_label}
% it will automatically put Appendix, Figure, Section, etc before the reference
\usepackage[capitalise]{cleveref}

@@

At the lowest level, we have two types of MPC instructions 1) local/non-interactive instruction (i.e. \add) and 2) remote/interactive instruction (i.e. \mul). Let the cost of a single \add instruction be given by a monotonically decreasing $f(n)$, where the argument $n$ is the number of \add instructions being executed in parallel. Similarly the cost of a single \mul is given by a monotonically decreasing function $g(n)$.

Given a serial schedule (a linear graph) of an MPC program i.e. a sequence of instructions $\mathcal{S}:=(S_1, \dots, S_n)$, where $S_i \in \{\text{\add, \mul}\}, 1 \leq i \leq n$, and a def-use dependency graph $G(V, E)$ corresponding to $\mathcal{S}$, our task is to construct a parallel schedule (another linear graph) $\mathcal{P} := (P_1, \dots, P_n)$ observing the following conditions:

\begin{enumerate}
    \item Multiple, not necessarily continuous instructions of the same type (i.e. either \add or \mul) from $\mathcal{S}$ can be grouped into a single $P_i$  However, all such instructions must be of the same type (either \add or \mul).
    \item Def-use dependencies of the graph $G(V, E)$ are should be preserved i.e. if instructions $S_i, S_j, i < j$ are a def-use (an edge exists from $S_i$ to $S_j$ in $G$), then they can only be mapped to $P_{i'}, P_{j'}, i' < j'$.
\end{enumerate}

Our goal is to construct minimize the height of the graph $\mathcal{P}$. Indeed, a graph $\mathcal{P}$ with minimum height will maximize parallelization.

\paragraph{Correctness} Correctness of $\mathcal{P}$ is guaranteed by definition. Since def-use dependencies are preserved, the function (being computed) remains the same.

\paragraph{Cost Comparison} For the sequential schedule $\mathcal{S}$ consisting of $L$ local and $R$ remote instructions, the total cost is $\mathit{cost}(\mathcal{S}) = L \cdot f(1) + R \cdot g(1)$. In the extreme case where all $L$ and all $R$ instructions can be parallelized, the cost of $\mathcal{P}$ is $\mathit{cost}(\mathcal{P}) = L \cdot f(L) + R \cdot g(R)$. Since both $f$ and $g$ are monotonically decreasing, $\mathit{cost}(\mathcal{P}) < \mathit{cost}(\mathcal{S})$. Cost of all other parallel schedules lies between the extremes of $\mathit{cost}(\mathcal{S})$ and $\mathit{cost}(\mathcal{P})$.

Note that we use an MPC-Source control flow graph (CFG) $G'(V', E')$ along with def-use graph $G(V,E)$ to construct $\mathcal{P}$. We consider a linearized MPC schedule $\mathcal{S}$ above for ease of exposition. The argument becomes slightly more involved when dealing with a graph $G'$ that may contain cycles.

% Given a def-use dependency graph $G(V, E)$ that is constructed from for the input program. Each vertex/node in this graph represents an operation/instruction in the program and an edge represents dependence i.e. an edge from $V_1$ to $V_2$ means the result of operation in $V_1$ is used in the operation in $V_2$. To build a parallel schedule, we translate this graph $G(V, E)$ into another dependency graph $G'(V', E')$ such that:

% \begin{enumerate}
%     \item We collapse multiple vertices of the same type (i.e. same operation) in $V$ into a single vertex in $V'$.
%     \item dependencies are not broken i.e. If there is an edge from $V_1$ to $V_2$ ($V_1, V_2 \in V$), and we put them in vertices $V'_1, V'_2 \in G'$ respectively, then an edge must exist from $V'_1$ to $V'_2$.
% \end{enumerate}

% We want to build a graph $G'$ with fewest possible vertices since fewer vertices means more parallelization.

\ishaq{TODO: @Ana: I am still confused about what prevents a reviewer from saying, "you argue correctness of the schedule from unrolled MPC, you don't argue it for the pre-unrolled CFG that you use". Please comment.}

@@

\ana{Ishaq?}

\ishaq{The follow text is all from Linsay's thesis}.
\brevised{
We consider two operations, call them $A$ and $M$. $A$ and $M$ are two abstract MPC instruction, 
but as an example, $A$ stands for the ADD MPC instruction 
and $M$ stands for the MUL instruction. Each instruction in the program is either an $A$-instruction or
an $M$-instruction. In order to benefit from parallelization/amortization, we must schedule two or more 
$A$-instructions in the same parallel node (or two or more $M$-instructions in the same parallel node). 
Scheduling $A$-instructions in parallel with $M$-instruction does not benefit from amortization.
It incurs the exact same cost as scheduling the $A$-instructions in a node $P_A$, scheduling the $M$-instructions 
in a node $P_M$, and having $P_A$ precede $P_M$ in the parallel schedule. This is the difference 
between classical scheduling, as studied in parallel computing, and MPC scheduling.

In our theoretical treatment, we make the following assumptions:

\item $A$ and $M$ are of equal cost, 1 unit.
\item There is unlimited bandwidth---i.e., a single $A$-instruction (or $M$-instruction) costs as much as $N$ amortized $A$-instructions 
(or $M$-instructions), namely 1 unit.

%  Furthermore the run time cost of 
% computing operations of the same type remains the same regardless of the number of operations that are run.
% Roughly speaking, 1 A operation costs the same as 100 A operations and 1000 A operations.

Consider a loop body that consists of $n$ sequences: $S_1$, ... $S_n$ of $A$ and $M$ instructions. 
More precisely, the loop body is such that its instructions can be grouped into such sequences. 
$S_1$, ... $S_n$ can execute in parallel, however, all instructions within a sequence must 
execute sequentially. For example, consider the three sequences (the right arrow indicates a \emph{dependence},
meaning that the source node must execute before the target node): 
\begin{enumerate}
\item $A \rightarrow M \rightarrow A$
\item $A \rightarrow A \rightarrow A$
\item $M \rightarrow A \rightarrow M$
\end{enumerate}

A \emph{schedule} $P: P_1 \rightarrow P_2 \dots \rightarrow P_k$ is such that for each sequence 
$S_i$ in the set, if $S_i[k]$ precedes $S_i[k']$ in $S_i$ then $S_i[k]$ is scheduled in node $P_l$, $S_i[k]$ 
is scheduled in node $P_{l'}$, and $P_l$ precedes $P_{l'}$ in $P$.

The cost of a schedule $P$ is

\begin{equation}
\mathit{cost}(P) = \sum_{i=1}^k \mathit{cost}(P_i)
\end{equation}

where $\mathit{cost}(P_i) = 1$ if $P_i$ consists of $A$-instructions only, or $M$-instructions only, 
and $\mathit{cost}(P_i) = 2$ if $P_i$ mixes $A$-instructions and $M$-instructions.

The problem is to find a schedule $P$ with \emph{minimal cost}. For example, 
a schedule with minimal cost for the sequences above is
\[ A(1), A(2) \rightarrow M(1), A(2), M(3) \rightarrow A(1), A(2), A(3) \rightarrow M(3) \]
(The parentheses above indicate the sequence where the instruction comes from: (1), (2), or (3).)
The cost of this schedule is 5.

The problem of finding a schedule $P$ with a minimal $cost(P)$ for a given loop body has been shown
to be an NP-Hard problem, as it can be reduced to the problem of finding a \emph{shortest common supersequence}, 
a known NP-Hard problem\cite{Maier1978},\cite{Vazirani2010}. The shortest common supersequence problem is as follows: 
given two or more sequences find the the shortest sequence that contains all of the original sequences. This can be solved
in $O(n^k)$ time, where $n$ is the cardinality of the longest sequence and $k$ is the 
number of sequences. For our problem $n$ is the maximum length of a node and $k$ is the number
of total number of nodes. %\ana{I think it is clear without additional explanation...}
%\ana{Also, add the full statement of the problem here. Now it's not clear what's NP-hard, what's n, what's k, etc.?}
%\lindsey{not exactly sure what need to go it there, but does that make it clearer?}

To see the reduction, 
suppose $P$ is a schedule with minimal cost (computed by a black-box algorithm). 
We can derive a schedule $P'$ with the same cost as $P$, by mapping each mixed node $P_i \in P$ 
to two consecutive nodes in $P'$: an $A$-instruction node followed by an $M$-instruction node.
Clearly, $P'$, which now is a sequence of $A$'s and $M$'s, is a supersequence of each sequence 
$S_i$, i.e., $P'$ is a common supersequence 
of $S_1 \dots S_n$. It is also a shortest common supersequence. (To see this, suppose, a 
shorter common supersequence, $P''$, exists. $P''$ is a schedule of $S_1 \dots S_n$
and the $\mathit{cost}(P'')$ equals the length of $P''$. Since $P'$ is longer than
$P''$, and $\mathit{cost}(P') = \mathit{cost}(P)$, that means $\mathit{cost}(P'') < \mathit{cost}(P)$, 
which is a contradiction since $P$ is a schedule with minimal cost.)
}

@@

\COMMENT { Phase 3: Dimensionality reduction: removing unnecessary dimensionality. }

@@

\subsection{Towards Extension of Basic Vectorization}

@@

\begin{algorithmic}
\STATE for i in range(N):
\STATE \hspace{0.25cm} A[i] = B[i] + 10;
\STATE \hspace{0.25cm} B[i] = A[i] * D[i-1];
\STATE \hspace{0.25cm} C[i] = A[i] * D[i-1];
\STATE \hspace{0.25cm} D[i] = B[i] * C[i];
\end{algorithmic}

@@

\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{A}_0$ = $\phi(\texttt{A},\texttt{A}_1)$
\STATE 2. $\texttt{B}_0$ = $\phi(\texttt{B},\texttt{B}_1)$
\STATE 3. $\texttt{C}_0$ = $\phi(\texttt{C},\texttt{C}_1)$
\STATE 4. $\texttt{D}_0$ = $\phi(\texttt{D},\texttt{D}_1)$
\STATE 5. $\texttt{A}_1$ = $\texttt{A}_0$; $\texttt{A}_1$[i] = $\texttt{B}_0$[i] + 10; \COMMENT{equiv. to Update}
\STATE 6. $\texttt{B}_1$ = $\texttt{B}_0$; $\texttt{B}_1$[i] = $\texttt{A}_1$[i] * $\texttt{D}_0$[i-1];
\STATE 7. $\texttt{C}_1$ = $\texttt{C}_0$; $\texttt{C}_1$[i] = $\texttt{A}_1$[i] * $\texttt{D}_0$[i-1];
\STATE 8. $\texttt{D}_1$ = $\texttt{D}_0$; $\texttt{D}_1$[i] = $\texttt{B}_1$[i] * $\texttt{C}_1$[i];
\end{algorithmic}

There is a cycle around $\texttt{B}_0$ = $\phi(\texttt{B},\texttt{B}_1)$ that includes statement $\texttt{A}_1$[i] = $\texttt{B}_0$[i] + 10;
and that statement won't be vectorized even though in fact there is no loop-carried dependency from the write of $\texttt{B}_1$[i] at 8 to the
read of $ ... = \texttt{B}_0$[i] at 6.

@@

Consider the earlier example. There is a single loop, $i$. Clearly, there is no pair $\underline{i}$ and $\underline{i}'$, where $\underline{i} < \underline{i}'$ that make $\underline{i} = \underline{i}'$ (due to the def-use pairs of \texttt{A} 6-8 and 6-10).
Therefore, we remove the back edge from 6 to 1. Analogously, we remove the back edges from 8 to 2 and 10 to 3. However, there are many values $\underline{i} < \underline{i'}$ that make $\underline{i} = \underline{i'}-1$ and the back edge from 12 to 4 remains (def-use pairs for \texttt{D}). As a result of removing these spurious edges, Basic Vectorization will find that statement 6 is vectorizable. Statements 8, 10 and 12 will correctly appear in the FOR loop.

@@

\ana{TODO: I think we should implement this.}

@@

\subsection{Extension of Basic Vectorization with Array Writes}

@@

For now, we restrict array updates to \emph{canonical updates}. Assume (for simplicity) a two-dimensional array $\texttt{A}[I,J]$. A canonical update is the following:
\begin{algorithmic}
\STATE for i in range(I):
\STATE \hspace{0.25cm} for j in range(J):
\STATE \hspace{0.55cm} ...
\STATE \hspace{0.55cm} A[i,j] = ...
\STATE \hspace{0.55cm} ...
\end{algorithmic}

The update $\texttt{A}[i,j]$ can be nested into an inner loop and there may be multiple updates, i.e., writes to $\texttt{A}[i,j]$. However, update such as $\texttt{A}[i-1,j] = ...$ or $\texttt{A}[i-1,j-1] = ...$, etc., is not allowed. Additionally, while there could be several different loops that perform canonical updates, they must be of the same dimensionality, i.e., an update of higher or lower dimension, e.g., $A[i,j,k] = ...$ is not allowed. We compute the \emph{canonical dimensionality} of each write array in the obvious way \emph{before basic vectorization}. This restriction simplifies reasoning in this early stage of the compiler; we will look to relax the restriction in future work.

@@

Reads through an arbitrary formula, such as $\texttt{A}[i-1]$ for example, are allowed; currently, our projection function returns dummy values if the read formula is out of bounds; we assume the programmer ensures that the program still computes correct output in this case.

@@

There are two changes to Basic Vectorization to account for write arrays. We can account for them during processing of raise dimension/drop dimension in Phase 1 of Basic Vectorization.

One change to Basic vectorization is the expansion of dimension if the array write or read occurs in a nested loop. That is, if there is an update $\texttt{A}[i,j] = ...$ that occurs in loop dimensionality $i,j,k$, $\texttt{A}[i,j]$ will be rewritten into $\texttt{A}[i,j,k]$. Similarly, a read $\texttt{A}[f(i,j)]$ will be rewritten into $\texttt{A}[f(i,j),k]$. ($\texttt{A}[f(i,j),k]$ will be reshaped as $f(i,j)*I*J+k$ during Phase 1 of Basic Vectorization. \ana{TODO: Double check.})

The other change concerns def-use edges $X \rightarrow Y$ where $X$ defines and $Y$ uses an array variable. The definition can be an update $A_2 = update(A_1,...)$ or a pseudo $\phi$-node $A_2 = \phi(A_0,A_1)$. Note that $\phi$ nodes for arrays are no subscript operations on those $\phi$-nodes the way there are in scalar arrays. These edges are not handled in the same way as in Basic Vectorization, specifically, we do not raise and drop dimension as we do for scalars in Basic Vectorization. A key invariant is that the dimension of an array $A$ cannot go above or drop below $A$'s canonical dimensionality. We enumerate the cases of def-use edges.

@@

First, the canonical dimensionality of all \texttt{A,B,C} and \texttt{D} is 1. 
Thus, there is no addition of extra dimension for inner loops. After Phase 1 
of Basic Vectorization the Aiken's array write example will be (roughly) as follows:

\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{A}_1$ = $\phi(\texttt{A}_0,\texttt{A}_2)$
\STATE 2. $\texttt{B}_1$ = $\phi(\texttt{B}_0,\texttt{B}_2)$
\STATE 3. $\texttt{C}_1$ = $\phi(\texttt{C}_0,\texttt{C}_2)$
\STATE 4. $\texttt{D}_1$ = $\phi(\texttt{D}_0,\texttt{D}_2)$
\STATE 5. $\texttt{A}_2 = update(\texttt{A}_1, \stackrel{\rightarrow}{i}, \texttt{B}_1[\stackrel{\rightarrow}{i}] + 10)$;
\STATE 6. $\texttt{B}_2 = update(\texttt{B}_1, \stackrel{\rightarrow}{i}, \texttt{A}_2[\stackrel{\rightarrow}{i}] * \texttt{D}_1[\stackrel{\rightarrow}{i-1}])$;
\STATE 7. $\texttt{C}_2 = update(\texttt{C}_1, \stackrel{\rightarrow}{i}, \texttt{A}_2[\stackrel{\rightarrow}{i}] * \texttt{D}_1[\stackrel{\rightarrow}{i-1}])$;
\STATE 8. $\texttt{D}_2 = update(\texttt{D}_1; \stackrel{\rightarrow}{i}, \texttt{B}_2[\stackrel{\rightarrow}{i}] * \texttt{C}_2[\stackrel{\rightarrow}{i}])$;
\end{algorithmic}

Phase 2 computes the closure of 4; $cl = \{4,6,7, 8\}$ 
while 5 is vectorizable. Recall that 1,2, and 3 are targetless phi-nodes. Since the closure $cl$ includes updates to $B$ and $C$,
the corresponding phi-nodes are added to the closure, i.e., the FOR loop. The uses of $A_1$ and $B_1$ in the vectorized statement
turn into uses of $A_0$ and $B_0$ respectively. (But note that $A_0$ is irrelevant; the updates writes into array $A_2$ in parallel.)

\begin{algorithmic}
\STATE 1. $\texttt{A}_2 = update(\texttt{A}_0; \stackrel{\rightarrow}{i}, \mathit{ADD\_SIMD}(\texttt{B}_0[\stackrel{\rightarrow}i],[10,...])$ \COMMENT{Fully vectorized, size N.}
\STATE FOR i=0; i<N; i++; \COMMENT{ MOTION loop }
\STATE 2. $\texttt{B}_1$ = $\phi(\texttt{B}_0,\texttt{B}_2)$
\STATE 3. $\texttt{C}_1$ = $\phi(\texttt{C}_0,\texttt{C}_2)$
\STATE 4. $\texttt{D}_1$ = $\phi(\texttt{D}_0,\texttt{D}_2)$
\STATE 5. $\texttt{B}_2 = update(\texttt{B}_1, i, \mathit{MUL}(\texttt{A}_2[i], \texttt{D}_1[{i-1}])$
\STATE 6. $\texttt{C}_2 = update(\texttt{C}_1, i, \mathit{MUL}(\texttt{A}_2[i], \texttt{D}_1[{i-1}])$
\STATE 7. $\texttt{D}_2 = update(\texttt{D}_1, i, \mathit{MUL}(\texttt{B}_2[i], \texttt{C}_1[i])$
\end{algorithmic}

\ana{TODO: Format either Histogram or Matrix Multiplication. Both are interesting examples that use output arrays.}
